JuPOETs: A Constrained Multiobjective Optimization Approach to Estimate Biochemical Model Ensembles in the Julia Programming Language

### 2016-11-21 response_to_reviewers/Paper_v2.tex
Submitted version with changes

JuPOETs integrates simulated annealing (SA) \cite{Kirkpatrick:1983aa} with Pareto ranking to estimate parameter sets on or near the optimal tradeoff surface between competing objectives (Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
A tradeoff surface defines the best possible performance for every conflicting objective, such that an increase in the performance of one objective does not decrease the performance of at least one other objective.  Pareto rank is a measure of distance away from the optimal tradeoff surface.
Thus, the central idea underlying POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (parameter guess at iteration $i+1$) and Pareto rank (Fig. \ref{fig:fig-MOSAalgorithm}).

@@

The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the error at iteration $i+1$ to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i$ meeting a ranking criterion). Pareto rank is a measure of distance from the trade-off surface;
parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).

@@

where $T$ is the computational annealing temperature that provides control over how strictly Pareto rank is enforced.

@@

However, as $T$ is reduced, the probability of accepting a high-rank set decreases.

@@

To use JuPOETs, the user specifies the \texttt{neighbor}, \texttt{acceptance}, \texttt{cooling} and \texttt{objective} functions along with an initial decision variable guess.
If the user is operating JuPOETs in hybrid mode, then a \texttt{refinement} function pointer must also be specified; the default \texttt{refinement} implementation is simply a pass through function. Default implementations of the \texttt{neighbor}, \texttt{acceptance} and \texttt{cooling} functions can be used directly, or they can be overridden by user defined logic. However, the user must provide an implementation of the \texttt{objective} function and provide an initial decision variable guess.
In addition, there are several user configurable parameters that can be adjusted to control the performance of JuPOETs:
\texttt{maximum\_number\_of\_iterations} controls the number of iterations per temperature (default 20);
\texttt{rank\_cutoff} controls the upper rank bound on the solution archive (default 5);
\texttt{temperature\_min} controls the minimum temperature after which JuPOETs returns the error and solution archives (default 0.001);
\texttt{show\_trace} controls the level of output shown to the user (default true).
After the completion of the run, JuPOETs returns the parameter solution archive $\mathcal{S}$, objective archive $\mathcal{O}$ and rank archive $\mathcal{R}$.
The parameter solution archive $\mathcal{S}$ contains is an $\mathcal{D}\times\mathcal{A}$ array, where $\mathcal{A}$ denotes the number of solutions in the archive when
JuPOETs terminated. On the other hand, the objective archive $\mathcal{O}$ is an $\mathcal{K}\times\mathcal{A}$ array containing the performance values for each objective corresponding the columns of $\mathcal{S}$. Lastly, JuPOETs returns the rank archive $\mathcal{R}$ which is an $\mathcal{A}\times{1}$ array of Pareto ranks corresponding to the
columns of $\mathcal{S}$.

@@

>>>>

JuPOETs integrates simulated annealing (SA) \cite{Kirkpatrick:1983aa} with Pareto ranking to estimate decision variables on or near the optimal tradeoff surface between competing objectives (Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
A tradeoff surface defines the best possible performance for every conflicting objective, such that an increase in the performance of one objective does not decrease the performance of at least one other objective.
Pareto rank is a scalar measure of distance away from the optimal tradeoff surface (low rank is near the surface, while higher ranks are progressively further away).
Thus, the central idea underlying POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (decision variable guess at iteration $i+1$) and the scalar Pareto rank (Fig. \ref{fig:fig-MOSAalgorithm}). Traditional simulated annealing uses a scalar performance value e.g., simulation error to make a probabilistic decision to keep or reject a set of decision variables; decision variables with better performance are always accepted, while those with worse performance are sometimes accepted depending upon a parameter called the temperature. On the other hand, JuPOETs makes this same decision using the Pareto rank instead of a single performance objective.
The problem of estimating biochemical model parameters from experimental data is typically posed as an error minimization problem over continuous real-valued decision variables (model parameters) subject to the model equations.

@@

The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the error at iteration $i+1$ to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i$ meeting a ranking criterion).
Parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).

@@

where $T$ is the simulated annealing temperature; the temperature provides control over how strictly decreasing Pareto rank is enforced.

@@

However, as $T$ is reduced as a function of iteration count (using the \texttt{cooling} function), the probability of accepting a high-rank set decreases.

@@

To use JuPOETs, the user specifies the \texttt{neighbor}, \texttt{acceptance}, \texttt{cooling} and \texttt{objective} functions along with an initial decision variable guess.
Default implementations of the \texttt{neighbor}, \texttt{acceptance} and \texttt{cooling} functions can be used directly, or they can be overridden by user defined logic.
However, the user must provide an implementation of the \texttt{objective} function and provide an initial decision variable guess.
Lastly, if the user is operating JuPOETs in hybrid mode, then a \texttt{refinement} function pointer must also be specified.
Hybrid mode temporarily switches the search from a multiobjective to a single objective problem, where the sum of the objective functions can be used to update the best (or initial) parameter guess.
The specific hybrid mode search logic is up to the user; by default hybrid mode is off, and the default \texttt{refinement} implementation is simply a pass through function.
However, we have shown previously that POETs operated in hybrid mode (where the single objective problem used a pattern search approach)
had better performance that POETs alone \cite{Song:2010aa}. Thus, hybrid mode is generally recommended for most applications.
In addition, there are several user configurable parameters that can be adjusted to control the performance of JuPOETs:
\texttt{maximum\_number\_of\_iterations} controls the number of iterations per temperature (default 20);
\texttt{rank\_cutoff} controls the upper rank bound on the solution archive (default 5);
\texttt{temperature\_min} controls the minimum temperature after which JuPOETs returns the error and solution archives (default 0.001);
\texttt{show\_trace} controls the level of output shown to the user (default true).
After the completion of the run, JuPOETs returns the parameter solution archive $\mathcal{S}$, objective archive $\mathcal{O}$ and rank archive $\mathcal{R}$.
The parameter solution archive $\mathcal{S}$ contains is an $\mathcal{D}\times\mathcal{A}$ array, where $\mathcal{A}$ denotes the number of solutions in the archive when
JuPOETs terminated. On the other hand, the objective archive $\mathcal{O}$ is an $\mathcal{K}\times\mathcal{A}$ array containing the performance values for each objective corresponding the columns of $\mathcal{S}$. Lastly, JuPOETs returns the rank archive $\mathcal{R}$ which is an $\mathcal{A}\times{1}$ array of Pareto ranks corresponding to the
columns of $\mathcal{S}$. One technical note, if JuPOETs is run from multiple starting locations, and the archives from each of these runs is combined into a single collective archive, the combined parameter rank archive may become invalid. In these cases, it is required to re-rank the parameter sets using the built-in \texttt{rank} function to produce a collective parameter ranking.

@@

Currently, JuPOETs does not consider parameter identifiability when constructing parameter ensembles.
Although JuPOETs produces parameter estimates that give model performance similar to the training data,
we do not have strict statistical confidence that the \textit{true} parameter values are contained within the ensemble.
However, despite this, ensembles produced by POETs can be predictive \cite{Lequieu:2011aa,Song:2009aa}.
Thus, JuPOETs produces a collection of parameters that are constrained by the performance of the model,
and not by specific hypotheses regarding the individual values of the raw model parameters.
Of course, knowledge of specific parameter values, or the relationship between parameter combinations, can be used to inform the search through either bounds or problem specific constraints (for example, as demonstrated in the first example problem.)
